---
title: "EDS 222: Homework 1"
date: "Due 10/14/24"
author: "Liz Peterson"
editor_options: 
  chunk_output_type: console
---

```{r}
# load in packages
library(here)
library(tidyverse)
```

## Background

*(The case study in this exercise is based on reality, but does not include actual observational data.)*

In this exercise we will look at a case study concerning air quality in South Asia. The World Health Organization estimates that air pollution kills an estimated seven million people per year, due to its effects on the cardiovascular and respiratory systems. Out of the 40 most polluted cities in the world, South Asia is home to 37, and Pakistan was ranked to contain the second most air pollution in the world in 2020 (IQAIR, 2020). In 2019, Lahore, Pakistan was the 12th most polluted city in the world, exposing a population of 11.1 million people to increased mortality and morbidity risks.

In this exercise, you are given two datasets from Lahore, Pakistan and are asked to compare the two different data collection strategies from this city. These data are:

-   Crowd-sourced data from air quality monitors located in people's homes. These data are voluntarily collected by individual households who choose to install a monitor in their home and upload their data for public access.

-   Official government data from monitors installed by government officials at selected locations across Lahore. There have been reports that government officials strategically locate monitors in locations with cleaner air in order to mitigate domestic and international pressure to clean up the air.

::: callout-note
All data for EDS 222 will be stored on the Taylor server, in the shared `/courses/eds-222/data/` directory. Please see material from EDS 214 on how to access and retrieve data from Taylor. These data are small; all compute can be handled locally. Thanks to Bren PhD student Fatiq Nadeem for assembling these data!
:::

In answering the following questions, please consider the lecture content from class on sampling strategies, as well as the material in Chapter 2 of [*Introduction to Modern Statistics*](https://openintro-ims.netlify.app/data-design). Include in your submission your version of this file "`eds-222-hw1.qmd`" and the rendered HTML output, each containing complete answers to all questions *as well as the associated code*. Questions with answers unsupported by the code will be marked incomplete. Showing your work this way will help you develop the habit of creating reproducible code.

## Assessment

### Question 1

Load the data from each source and label it as `crowdsourced` and `govt` accordingly. For example:

``` r
crowdsourced <- readRDS(file.path("data", "airpol-PK-crowdsourced.RDS"))
govt <- readRDS(file.path("data", "airpol-PK-govt.RDS"))
```

```{r}
crowdsourced <- readRDS(here("data", "airpol-PK-crowdsourced.RDS"))
govt <- readRDS(here("data", "airpol-PK-govt.RDS"))
```

::: callout-warning
There's an implicit assumption about file organization in the code above. What is it? How can you make the code work?
:::

The file organization above assumes that when we downloaded the data from cyberduck we put it in a data folder that we created and put in our repository. I made the code work by creating a data folder, and I used the here library to navigate to that folder in my current working directory.

1.  These dataframes have one row per pollution observation. How many pollution records are in each dataset?

```{r}
nrow(crowdsourced)
nrow(govt)
```

There are 5488 pollution records in the crowd-sourced dataframe, and there are 1960 pollution records in the govt dataframe.

2.  Each monitor is located at a unique latitude and longitude location. How many unique monitors are in each dataset?

```{r}
unique_monitors_crowd <- crowdsourced %>%
  group_by(longitude, latitude) %>%
  summarise(n = cur_group_id())

unique_monitors_govt <- govt %>%
  group_by(longitude, latitude) %>%
  summarise(n = cur_group_id())
 
print(nrow(unique_monitors_crowd))
print(nrow(unique_monitors_govt))
```

::: callout-tip
`group_by(longitude,latitude)` and `cur_group_id()` in `dplyr` will help in creating a unique identifier for each (longitude, latitude) pair.
:::

The crowd-sourced data has 14 unique monitors, and the government data has 5 unique monitors.

### Question 2

The goal of pollution monitoring in Lahore is to measure the average pollution conditions across the city.

1.  What is the *population* in this setting? Please be precise.

    The population in this Lahore would be the entire city's pollution levels, regardless of location. Additionally, the population is specifically the data from November 2018 to November 2019.

2.  What are the *samples* in this setting? Please be precise.

    The samples in this setting are the crowd-sourced data and the government sourced data.

3.  These samples were not randomly collected from across locations in Lahore. Given the sampling approaches described above, discuss possible biases that may enter when we use these samples to construct estimates of population parameters.

    There are two pitfalls for possible bias in these sampling locations. First, the people who volunteer to have these monitors in their homes are likely people who would be willing to work to lower their pollution. Additionally, the government decided to place monitors in locations that had lower pollution levels to make themselves look better.

### Question 3

1.  For both the government data and the crowd-sourced data, report the sample mean, sample minimum, and sample maximum value of PM 2.5 (measured in $\mu g/m^3$).

    ```{r}
    mean(crowdsourced$PM)
    min(crowdsourced$PM)
    max(crowdsourced$PM)

    mean(govt$PM)
    min(govt$PM)
    max(govt$PM)
    ```

    For the crowd-sourced data, the sample mean is 70.2008 $\mu g/m^3$, the sample minimum is 20 $\mu g/m^3$, and the sample maximum is 120 $\mu g/m^3$. For the government data, the sample mean is 39.64694 $\mu g/m^3$, the sample minimum is 15 $\mu g/m^3$, and the sample maximum is 65 $\mu g/m^3$.

2.  Discuss any key differences that you see between these two samples.

    The mean for the crowd-sourced data is much higher, which makes sense considering the maximum for the crowd-sourced data is almost double the maximum for the government data.

3.  Are the differences in mean pollution as expected, given what we know about the sampling strategies?

    The differences are as expected when we look at the sampling strategies. The government reportedly placed monitors in areas with cleaner air in order to "mitigate domestic and international pressure to clean up the air." Additionally, there is bias in the crowd-sourced data because we can assume that the kind of person who would volunteer to have this monitor placed near them might already have some more consciousness about pollution.

### Question 4

Use the location of the air pollution stations for both of the sampling strategies to generate a map showing locations of each observation. Color the two samples with different colors to highlight how each sample obtains measurements from different parts of the city.

::: callout-tip
`longitude` indicates location in the *x*-direction, while `latitude` indicates location in the *y*-direction. With `ggplot2` this should be nothing fancy. We'll do more spatial data in `R` later in the course.
:::

```{r}
ggplot() +
  geom_point(data = crowdsourced,
             aes(x = longitude, y = latitude, color = "Crowd-sourced data")) +
  geom_point(data = govt,
             aes(x = longitude, y = latitude, color = "Government data")) +
  scale_color_manual("Monitor type", values = c("Crowd-sourced data" = "red", "Government data" = "blue")) +
  labs(title = "Location of observations",
       x = "Longitude",
       y = "Latitude")
```

# use geom_point here

### Question 5

The local newspaper in Pakistan, *Dawn*, claims that the government is misreporting the air pollution levels in Lahore. Do the locations of monitors in question 4, relative to crowd-sourced monitors, suggest anything about a possible political bias?

There could certainly be a political bias present, considering the way that the government placed their monitors to avoid having to actually solve any pollution problems. This could favor those not actually afflicted by the rising pollution levels.

### Question 6

Given the recent corruption in air quality reporting, the Prime Minister of Pakistan has hired an independent body of environmental data scientists to create an unbiased estimate of the mean PM 2.5 across Lahore using some combination of both government stations and crowd sourced observations.

NASA's satellite data indicates that the average PM across Lahore is 89.2 $\mu g/m^3$. Since this is the most objective estimate of population-level PM 2.5 available, your goal is to match this mean as closely as possible by creating a new ground-level monitoring sample that draws on both the government and crowd-sourced samples.

#### Question 6.1

First, generate a *random sample* of size $n=1000$ air pollution records by (i) pooling observations across the government and the crowd-sourced data; and (ii) drawing observations at random from this pooled sample.

::: callout-tip
`bind_rows()` may be helpful.
:::

```{r}
# Add in column of to show where the data is coming from for 6.2
crowdsourced$kind <- "crowdsourced"
govt$kind <- "government"
pooled_df <- bind_rows(crowdsourced, govt)

rand_pooled_samp <- pooled_df %>%
  sample_n(1000)

dim(rand_pooled_samp)
```

Second, create a *stratified random sample*. Do so by (i) stratifying your pooled data-set into strata of 0.01 degrees of latitude, and (ii) randomly sampling 200 air pollution observations from each stratum.

```{r}
# round
round_pooled <- pooled_df %>%
  mutate(lat_round = round(latitude, digits = 2))

strata_pooled <- round_pooled %>%
  slice_sample(n = 200, replace = FALSE, by = lat_round)

print(dim(strata_pooled))

ggplot(strata_pooled) +
  geom_bar(aes(x = lat_round))
```

#### Question 6.2

Compare estimated means of PM 2.5 for each sampling strategy to the NASA estimate of 89.2 $\mu g/m^3$. Which sample seems to match the satellite data best? What would you recommend the Prime Minister do? Does your proposed sampling strategy rely more on government or on crowd-sourced data? Why might that be the case?

```{r}
mean_rand_samp <- mean(rand_pooled_samp$PM)

mean_strat_samp <- mean(strata_pooled$PM)

print(mean_rand_samp)
print(mean_strat_samp)
```

```{r}
strata_govt <- strata_pooled %>%
  filter(kind == "government")
strata_cs <- strata_pooled %>%
  filter(kind == "crowdsourced")

print(dim(strata_govt))
print(dim(strata_cs))
```

When doing a random sample of 1000 observations over a pooled data frame of both government sourced and crowd sourced data, the mean of PM 2.5 is 61.28. For a stratified random sample of the same, the mean of PM 2.5 is 67.31. This shows that the stratified random sample would be closer to the actual satellite data. Therefore, relying on this data would be better. Within the stratified sample, we see that 905 observations are coming from crowd sourced data monitors while only 95 are coming from those government monitors. This could be due to the bias discussed above.
